Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-05-11 documents/ms.tex
adding discussion of computation time and hierarchical inference

The main aim of this work was to develop a probabilistic rotation period
inference method.
Probabilistic periods are necessary for hierarchical Bayesian inference,
particularly when performing population analysis.
This new GP method is capable of generating a catalogue of probabilistic
rotation periods which could, for example, reveal the period distribution
(and therefore potentially the age distribution) of stars in the Milky Way.
There is still room for improvement since the uncertainty estimates are not
yet truly representative and it is still only an `effective' model, not a
physical one.
However, it is probabilistic and provides more accurate periods than
alternative methods.
{\bf Although GP inference is slow as it requires a `solve' and a matrix
determinant calculation at every step of an MCMC, a new GP solver, {\tt
celerite} \citep{Foremanmackey2017} shows promise for increasing the
computional speed of GP rotation period inference by several orders of
magnitude.
We intend to integrate {\tt celerite} into our code in the near future.}
The code used in this project is available at
\url{https://github.com/RuthAngus/GProtation/}.

>>>>

The main aim of this work was to develop a probabilistic rotation period
inference method.
Probabilistic periods are necessary for hierarchical Bayesian inference,
particularly when performing population analysis.
This new GP method is capable of generating a catalogue of probabilistic
rotation periods which could, for example, reveal the period distribution
(and therefore potentially the age distribution) of stars in the Milky Way.
There is still room for improvement since the uncertainty estimates are not
yet truly representative and it is still only an `effective' model, not a
physical one.
However, it is probabilistic and provides more accurate periods than
alternative methods.
{\bf Although GP inference is slow as it requires a `solve' and a matrix
determinant calculation at every step of an MCMC, a new GP solver, {\tt
celerite} \citep{Foremanmackey2017} shows promise for increasing the
computional speed of GP rotation period inference by several orders of
magnitude.
We intend to integrate {\tt celerite} into our code in the near future.
The issue of computational time is a personal one.
GP period inference is trivially parallelisable, so if one has access to
thousands of computer cores, one can infer rotation periods of thousands of
stars in a few hours (perhaps just a few minutes with {\tt celerite}).
For single stars GP rotation inference takes a reasonable amount of time and
has already been used on exoplanet hosts \citep[\eg][]{Haywood2014,
Vanderburg2015}.

Is it worth spending the extra computation power to get slightly more
accurate, probabilistic rotation periods for a large number of stars?
A growing number of astronomers are interested in hierarchical probabilistic
modelling, particularly in the stellar and exoplanet communities.
These individuals are mindful of carefully inferring probabilistic parameters
wherever possible.
Probabilistic inference also allows you to marginalise over, or ‘bypass’,
certain parameters.
For example, imagine trying to infer the distribution of rotation periods of a
sample of stars.
One could simply calculate rotation periods from the light curves and produce
a histogram of the rotation values.
This is `adding posteriors' which is not technically legal in probability
theory, you are only allowed to add (or sum) likelihoods.
In order to add likelihoods you have to jointly infer the rotation period of
each individual star and the parameters of the distribution, which you might
model as a Gaussian.
Here, the individual rotation periods of the stars are nuisance parameters:
you don’t care about their values, you only compute them in order to infer the
mean and variance of the period distribution.
Using hierarchical inference allows you to marginalise over individual
rotation periods and infer distribution parameters directly from the light
curves.
Many people are interested in the distribution of stellar ages across the
galaxy and rotation periods can be used as an age proxy via gyrochronology.
Probabilistic rotation period inference is therefore highly relevant to a
large number of galactic archaeology and exoplanet population studies which
require rotation period (or age) distributions.}

The code used in this project is available at
\url{https://github.com/RuthAngus/GProtation/}.
